---
title: "Georges Bank Rpath Model Notes"
output: html_document
bibliography: GBRpath.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(data.table); library(knitr)

```

Georges Bank
===

Georges Bank is a highly productive sub-marine plateau located directly east of Cape Cod, Massachusetts.  The bank has a long history of commercial fishing and is currently the focus of an example Fisheries Ecosystem Plan (eFEP) for the New England Fisheries Management Council.  This repository will house all of the code that goes into creating a mass balance food web model version of Georges Bank.  The model will be built using [Rpath](http://github.com/NOAA-EDAB/Rpath), the R implementaion of the mass balance algorithms. 

Rpath model structure
===
The geographic extent of the model will be the Georges Bank Ecological Production 
Unit developed by the Ecosystem Assessment and Dynamics Branch (EDAB) of the Northeast
Fisheries Science Center (NEFSC).

The base structure of the model is the EMAX model of @link_EMAX_2006.  However, it
will be much more disaggregated.  Here are the model nodes:

```{r model nodes}
groups <- data.table(Group = c('Seabirds', 'Seals', 'BalWhale', 'ToothWhale', 'HMS', 
                               'Sharks', 'AtlHerring', 'AtlMackerel', 'Butterfish', 
                               'SmPelagics', 'Mesopelagics', 'OtherPelagics', 'Cod', 
                               'Haddock', 'Goosefish', 'OffHake', 'SilverHake', 
                               'RedHake', 'WhiteHake', 'Redfish', 'Pollock', 'OceanPout', 
                               'BlackSeaBass', 'Bluefish', 'Scup', 'OtherDemersals', 
                               'SouthernDemersals', 'Fourspot', 'SummerFlounder', 
                               'AmPlaice', 'Windowpane', 'WinterFlounder', 'WitchFlounder', 
                               'YTFlounder', 'OtherFlatfish', 'SmFlatfishes', 'SpinyDogfish', 
                               'SmoothDogfish', 'Barndoor', 'WinterSkate', 'LittleSkate', 
                               'OtherSkates', 'Illex', 'Loligo', 'OtherCephalopods', 
                               'AmLobster', 'Macrobenthos', 'Megabenthos', 'AtlScallop', 
                               'Clams', 'OtherShrimps', 'Krill', 'Micronekton', 
                               'GelZooplankton', 'Mesozooplankton', 'Microzooplankton', 
                               'Phytoplankton'), 
                     type = c('Bird', 'Mammal', 'Whale', 'Whale', 'HMS', 'Shark',
                              rep('Forage', 6), rep('DemRound', 15), 
                              rep('DemFlat', 9), 'DemRound', 'DemRound', 
                              rep('Dem', 4), rep('PelInvert', 3), rep('BenInvert', 5),
                              rep('PelInvert', 3), rep('Zoop', 3), 'Phyto'))

kable(groups, caption = "Living nodes for Georges Bank Rpath model.", 
      booktabs = T, align = c('c', 'c'))
```

Several other groups showed up in the biomass data pull describe below but were 
not present for the whole time series.  Therefore they were omitted.

```{r extra nodes}
groups <- data.table(Group = c('AmShad', 'AtlHalibut', 'NShrimp', 'RedCrab', 'RiverHerring',
                               'StripedBass', 'Tilefish'),
                     Years = c(3, 2, 1, 2, 2, 2, 1))
kable(groups, caption = 'Species present during 2012-2016 on Georges Bank but not included.',
      booktabs = T, align = c('c', 'c'))
```

```{r other nodes}
groups <- data.table(Group = c('Detritus', 'Discards', 'DredgeScallop', 'DredgeClam', 
                               'Gillnet', 'Longline', 'Seine', 'PotTrap', 'OttertrawlSmall', 
                               'OttertrawlLarge', 'Midwater', 'OtherFisheries'),
                     Type = c(rep('Detrital Group', 2), rep('Fleet', 10)))
kable(groups, caption = 'Non-living nodes for the Georges Bank Rpath model.',
      booktabs = T, align = 'c')
```

Data pulls
===

## Biomass/ Biomass accumulation
For non-bivalve species that are encountered in our surveys I am using Bigelow Fall 
data 2012 to 2016 - without conversion factors. Fall 2017 
survey was conducted on the Pisces with no Mid coverage which will be an issue for 
future work...also spring/fall 2018 compromised due to ship issues.

I am using the scallop survey and clam survey for Atlantic Scallops and Ocean 
Quhogs/Surfclams respectively.  Note that the clam survey samples regions on a 
rotating basis.  See below table:
```{r clam table, echo = F, warning = F}
clam.tows <- data.table(Year = 2012:2016, ntows = c(0, 115, 29, 8, 166))
kable(clam.tows, caption = "Number of clam tows within the Georges Bank EPU.", 
      booktabs = T, align = c('c', 'c'))

```

Biomass for surveyed species is calculated as the mean from 2013 to 2015.  Clams 
will be an average of 2013 and 2016 as those two years are more representive than 
the others.

2012 through 2016 are used to see if there is a significant trend for a BA term.  Clams
were not assessed for a trend because there are only two data points.

Species not encountered by our survey are using EMAX values [@link_EMAX_2006].  I 
also assumed no BA term.

Three groups showed significant trends.  All three were positive which was a 
surprise. The three groups were Haddock, Southern Demersals, and Atlantic 
Mackerel.  Haddock and Southern Demersal make sense but not Mackerel.  An increasing
biomass would be counter to the resent assessmets.  So in talking with Sarah, I'm 
going to hold off on BA terms unless model doesn't balance right.

*Need to double check "Freshwater" designation*

## Landings/Discards

Pulling data from the stockefficiency tables for 2013 - 2015.  Landings will be 
an average of those years.  

Need to deal with skates which are landed as skates(ns).  Human consumption mainly
winter or thorny while bait mainly little skate (44th SAW).  Set all utilcd of 0 
to winter skates and utilcd of 7 to little skates.  End fate of observer data not 
known but only 7.6% of skates are listed as skates(ns).  The rest are identified
to species.  Therefore I'm just allowing the "OtherSkates" node absorb the unknown
skates.

Dividing otter trawl landings by large and small mesh.  The
[regulation](https://www.greateratlantic.fisheries.noaa.gov/regs/infodocs/small_mesh_exemption.pdf) 
for July 2015 was 6.5 inches.  Approximately 18% of otter trawl records (~11% by 
weight) do not have a mesh associated with them. 

Total landings by gear type are divided by the area of the NAFO Statistical Areas
that have been selected as Georges Bank.  This area is much larger than the survey
foot print.

Figuring out which otter trawl trips were small mesh was challenging.  Originally
I was using the otter trawl gear table (OBOTG).  Unfortunately it was not to the
haul level and some trips switched gear during the trip.  Some even switched between
small mesh and large mesh.  After extensive searching and frustration I founs the OBDBS
table OBOTGH which contains a link to both the haul (OBHAU) and gear tables (OBOTG).  The
species table (OBSPP) which contains the hail weights for both caught and discarded
species contains the primary key for hauls (LINK3).  OBOTGH has LINK3 and LINK4 
(Gear key) which can be used then to assign mesh size.  Not all catches had an 
associated gear entry but 95.8% did (which is better than the landings tables!). For
the purposes of calculating a DK ratio I am not worrying about the unassigned 5%.

Had to manually assign Rpath groups for the incidental takes in the observer 
database.  There were 16 sea turtles in the >500K records.  I decided this was
not enough to warrent their inclusion.  Also, incidental takes are measured as 
number of individuals not weight.  Estimate length is recorded but unfortunately
nearly 89% of records do not have on.  I did add animal condition to drop incidental
takes were the animal was still alive (~11% of records).

## PreBal

Coding up the PreBal routines from @link_prebal_2010.  Some of the checks are easy
to make general.  Some of them require knowing the classification of the species
node (i.e. demersal flatfish or small pelagic).  Will hard code for this project
and revisit at a later date.

Moving forward dynamically
===

## Notes on ABC (wikipedia)

ABC is approximate Bayesian computation - originally developed in genetics to estimate the posterior distribution when calculating the likelihood is impractical.

General reminder of Bayes Theorem:

$$  p(\theta | D) = \frac{p(D|\theta)p(\theta)}{p(D)} $$
where $p(\theta|D)$ is the posterior distribution, $p(D|\theta)$ the likelihood, $p(\theta)$ the prior distribution, and $p(D)$ the evidence or marginal likelihood.

All ABCs approximate the likelihood through simulations where outcomes are compared to observations.  The core is the ABC rejection algorithm where parameters are first sampled from the prior distribution. Given a sampled parameter point $\hat{\theta}$, a data set $\hat{D}$ is then simulated under the statistical model *M* specified by $\hat{\theta}$. If the generated $\hat{D}$ is too different from the observed data *D*, the sampled parameter value is discarded. The distance measure $\rho(\hat{D}, D)$ determines the level of discrepancy between \hat{D} and *D* based on a given metric (e.g. Euclidean distance). 

$$ \rho(\hat{D}, D) \leq \epsilon $$
The probability of generating a $\hat{D}$ that is within tolerance of *D* decreases as the dimensionality of the data increases.  To avoid this a set of lower-dimensional summary statistics, *S(D)* are used that captures the informatin in *D*.  The acceptance criteria then becomes:

$$ \rho(\hat{S(D)}, S(D)) \leq \epsilon $$ 

Package in R for doing approximate Bayesian computation: abc [@csillery_abc_2012]







References {#references .unnumbered}
===